---
title: "Mason Donehoo Project 1"
format: html
editor: visual
---

```{r}
#| include: FALSE
library(httr)
library(jsonlite)
library(dplyr)
library(tidyr)
library(tibble)
```

## Task 1: Data Processing

### Parsing Variable Values

For this section, I am retrieving and parsing metadata for the Public Use Microdata Sample (PUMS) Census API across four years: 2021, 2022, 2023, and 2024. 

The variables we are interested in are `AGEP`, `GASP`, `GRPIP`, `JWAP`, `JWDP`, `JWMNP`, `PWGTP`, `FER`, `HHL`, `SCH`, `SCHL`, `SEX`, `ST` (for 2021, 2022)/`STATE` (for 2023, 2024), `REGION`, and `DIVISION`. The variables for each year will be stored in a list that will then be saved in a `.rds` file under the `docs` folder.

```{r}
#Creating function to fetch and parse JSON variables for a single year
parse_year_metadata <- function(year){
  #Saves the url for whatever year the user is interested in
  url <- paste0("api.census.gov/data/", year, 
                "/acs/acs1/pums/variables.json")
  
  #Using GET() to contact the website and retrieve variable specifications
  response <- GET(url)
  
  #Parsing the JSON using fromJSON(rawToChar(...), simplifyVector = F). We use simplifyVector = F to ensure we return a list 
  parsed <- fromJSON(rawToChar(response$content), simplifyVector = F)
  
  #Extracting the `variables` element from the parsed data
  vars_list <- parsed$variables
  
  #Creating a vector of the variables we are interested in
  vars_needed <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
                   "FER", "SCH", "SCHL", "SEX", "ST", "STATE", "REGION", 
                   "Division")
  
  #Subsetting vars_list to get only the variables we are interested in from vars_needed
  vars_selected <- vars_list[names(vars_list) %in% vars_needed]
  
  #Returning the list of the variables with their values
  return(vars_selected)
}

#Saving the years we are interested in to a vector
years <- c(2021, 2022, 2023, 2024)

#Iterating across each year and saving the lists of the variables with their values
all_years <- setNames(lapply(years, parse_year_metadata), as.character(years))

#Creating data folder in my repo
if(!dir.exists("data")){
   dir.create("data")
  }

#Saving the lists in a .rds file under the data folder in my repo
saveRDS(all_years, file = "data/pums_variables.rds")
```

### Parsing API Data

This function focuses on querying the PUMS API and returning a formatted tibble. I will write a *helper* function to take what is returned by `GET()` and turn it into a formatted `tibble`. 

The *helper* function will be wrapped into a function that will query the API. The function will have the following default values: survey year `2024`, the numerical variables `AGEP` and `PWGTP`, the categorical variable `SEX`, and geography level `State`. 

The valid values for year are: 2021, 2022, 2023, and 2024 

For the numerical variables, `PWGTP` is always returned along with one other numerical variable: `AGEP`, `GASP`, `GRPIP`, `JWAP` (time), `JWDP` (time), and `JWMNP`. The function will turn variables into numeric values or time values when it is appropriate. For time values, a function is written to parse the time strings and returning the midpoint as a time value and NA for missing time values

For the categorical variables, the options for the user are: `FER`, `HHL`, `SCH`, `SCHL`, and `SEX` where `SEX` is the default as mentioned above. The function will turn variables into factors using the appropriate year's variable levels from my `.rds` file

For the geography level, the options for the user are: `Region`, `Division`, or `State` with `State` being the default as mentioned above. The "All" geography level is not included as the data being returned is too large. Furthermore, the user is only able to choose one of these geography levels to subset on for a given API call. If the user specifies multiple geography levels, specifies "All", or provides a value that is not `REGION`, `DIVISION`, or `STATE`, the function will return an error.

A default value for the `State`, `Region`, or `Division` being subset on is specified if the user does not specify one themselves. The allowable values for the geographies are in the `.rds` file that I created earlier, or the user can input an "all" option that would return all values of either the `State`, `Region`, or `Division`.

Finally, a formatted tibble will be returned


```{r}

```