---
title: "Mason Donehoo Project 1"
format: html
editor: visual
---

```{r}
#| include: FALSE
library(httr)
library(jsonlite)
library(dplyr)
library(tidyr)
library(tibble)
library(lubridate)
library(tidyverse)
```

## Task 1: Data Processing

### Parsing Variable Values

For this section, I am retrieving and parsing metadata for the Public Use Microdata Sample (PUMS) Census API across four years: 2021, 2022, 2023, and 2024. 

The variables we are interested in are `AGEP`, `GASP`, `GRPIP`, `JWAP`, `JWDP`, `JWMNP`, `PWGTP`, `FER`, `HHL`, `SCH`, `SCHL`, `SEX`, `ST` (for 2021, 2022)/`STATE` (for 2023, 2024), `REGION`, and `DIVISION`. The variables for each year will be stored in a list that will then be saved in a `.rds` file under the `docs` folder.

```{r}
#Creating function to fetch and parse JSON variables for a single year
parse_year_metadata <- function(year){
  #Saves the url for whatever year the user is interested in
  url <- paste0("api.census.gov/data/", year, 
                "/acs/acs1/pums/variables.json")
  
  #Using GET() to contact the website and retrieve variable specifications
  response <- GET(url)
  
  #Parsing the JSON using fromJSON(rawToChar(...), simplifyVector = F). We use simplifyVector = F to ensure we return a list 
  parsed <- fromJSON(rawToChar(response$content), simplifyVector = F)
  
  #Extracting the `variables` element from the parsed data
  vars_list <- parsed$variables
  
  #Creating a vector of the variables we are interested in
  vars_needed <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
                   "FER", "SCH", "SCHL", "SEX", "ST", "STATE", "REGION", 
                   "Division")
  
  #Subsetting vars_list to get only the variables we are interested in from vars_needed
  vars_selected <- vars_list[names(vars_list) %in% vars_needed]
  
  #Returning the list of the variables with their values
  return(vars_selected)
}

#Saving the years we are interested in to a vector
years <- c(2021, 2022, 2023, 2024)

#Iterating across each year and saving the lists of the variables with their values
all_years <- setNames(lapply(years, parse_year_metadata), as.character(years))

#Creating data folder in my repo
if(!dir.exists("data")){
   dir.create("data")
  }

#Saving the lists in a .rds file under the data folder in my repo
saveRDS(all_years, file = "data/pums_variables.rds")
```

### Parsing API Data

This function focuses on querying the PUMS API and returning a formatted tibble. I will first write a *helper* function to take what is returned by `GET()` and turn it into a formatted `tibble`. 

The *helper* function will be wrapped into a function that will query the API. The function will have the following default values: survey year `2024`, the numerical variables `AGEP` and `PWGTP`, the categorical variable `SEX`, and geography level `State`. 

The valid values for year are: 2021, 2022, 2023, and 2024 

For the numerical variables, `PWGTP` is always returned along with one other numerical variable: `AGEP`, `GASP`, `GRPIP`, `JWAP` (time), `JWDP` (time), and `JWMNP`. The function will turn variables into numeric values or time values when it is appropriate. For time values (i.e., `JWAP` and `JWDP`), a helper function is written to parse the time strings and return the midpoint as a time value and NA for missing time values

For the categorical variables, the options for the user are: `FER`, `HHL`, `SCH`, `SCHL`, and `SEX` where `SEX` is the default as mentioned above. The function will turn variables into factors using the appropriate year's variable levels from my `.rds` file

For the geography level, the options for the user are: `Region`, `Division`, or `State` with `State` being the default as mentioned above. The "All" geography level is not included as the data being returned is too large. Furthermore, the user is only able to choose one of these geography levels to subset on for a given API call. If the user specifies multiple geography levels, specifies "All", or provides a value that is not `Region`, `Division`, or `State`, the function will return an error.

A default value for the `State`, `Region`, or `Division` being subset on is specified if the user does not specify one themselves. The allowable values for the geographies are in the `.rds` file that I created earlier, or the user can input an "all" option that would return all values of either the `State`, `Region`, or `Division`.

Finally, a formatted tibble will be returned

Seeing as I know I'll need to create two helper functions, one to take what is returned by `GET()` and converting it into a formatted `tibble`, and one to parse time strings and return the midpoint as a time value, I will start by creating these helper functions. 

First I will create the first helper function that converts the `GET()` response object into a `tibble`
```{r}
#Creating function
parse_api_response <- function(response){
  #Parsing the JSON content from the API response object
  raw <- fromJSON(rawToChar(response$content))
  
  #Converting data rows into a tibble, dropping the first row as those are the variable names, and using .name_repair = "minimal" to make sure the tibble doesn't create it's own column names
  df <- as_tibble(raw[-1, , drop = F], .name_repair = "minimal")
  
  #Setting the column names to row 1
  colnames(df) <- raw[1, ]
  
  #Returning the tibble
  return(df)
}
```

Now I am creating the second helper function with the help of a LLM that will parse time strings if the user inputs either `JWAP` or `JWDP` into midpoints in minutes
```{r}
parse_time_string <- function(time_str){
  #Returns NA if the string is NA, if it is empty, or if it indicates a response that doesn't apply
  if(is.na(time_str) || time_str == "NA" || time_str == "" || 
     str_detect(time_str,"NA")){
    return(NA_real_)
  }
  
  #Extracts all matching HH:MM a.m./p.m. time strings from the text string
  time_clean <- gsub("\\.","",time_str)
  
  pattern <- "(\\d{1,2}):(\\d{2})\\s*([ap]\\.?m\\.?)"
  
  times <- str_extract_all(time_clean, pattern)
  
  parsed_time <- parse_date_time(times, order = "I:M p", quiet = T)
  
  mins <- hour(parsed_time) * 60 + minute(parsed_time)
    
  #Checking if two distinct time boundary endpoints were extracted
  if(length(mins) >= 2){
    start_min <- mins[1]
    end_min <- mins[2]
    
    midpoint <- (start_min + end_min) / 2
    
    return(midpoint %% (24*60))
  }else{
    return(mins[1])
  }
}
```

```{r}
parse_time_string2 <- function(x) {
  # 1. Guard check for NA, NULL, or non-workers
  if (is.null(x) || is.na(x) || grepl("^N/A", x, ignore.case = TRUE)) {
    return(NA_real_)
  }
  
  # 2. Extract hours, minutes, and am/pm markers
  # Matches "12:00 a.m.", "12:29 am", "12:00 AM", etc.
  pattern <- "(\\d{1,2}):(\\d{2})\\s*([ap]\\.?m\\.?)"
  matches <- stringr::str_match_all(tolower(x), pattern)[[1]]
  
  if (nrow(matches) == 0) {
    return(NA_real_)
  }
  
  # 3. Helper to convert extracted match row into minutes from midnight
  time_to_minutes <- function(match_row) {
    hr <- as.numeric(match_row[2])
    mn <- as.numeric(match_row[3])
    is_pm <- grepl("^p", match_row[4])
    
    if (hr == 12) {
      hr <- if (is_pm) 12 else 0
    } else if (is_pm) {
      hr <- hr + 12
    }
    
    return(hr * 60 + mn)
  }
  
  # 4. Calculate minutes from midnight for all times in the range
  mins <- apply(matches, 1, time_to_minutes)
  
  # 5. Compute range midpoint
  if (length(mins) >= 2) {
    start_min <- mins[1]
    end_min <- mins[2]
    
    # Handle midnight rollover (e.g., 11:50 p.m. to 12:10 a.m.)
    if (end_min < start_min) {
      end_min <- end_min + (24 * 60)
    }
    
    midpoint <- (start_min + end_min) / 2
    return(midpoint %% (24 * 60))
  } else {
    return(mins[1])
  }
}
```

Now that my two helper functions I will need have been created, I can write the function that will query the PUMS API and return a nicely formatted `tibble`. The default values I will use are mentioned above in the beginning of this section.
```{r}
get_census_tibble <- function(year = 2024,
                              numeric_vars = c("AGEP", "PWGTP"),
                              cat_vars = "SEX",
                              geo_level = "STATE",
                              geo_subset = NULL,
                              key = "9a9b66cf9a9433d327bf8c614f78a681d8df91b2"){
  
  #Checking to make sure the value specified for `year` is valid
  #Making sure the user only inputted one year
  if(length(year) != 1){
    #Produces error if more than one year is specified
    stop("Error: You may only use one value for 'year'")
  }
  
  #Making sure the user used a valid year i.e., 2021, 2022, 2023, 2024
  if(!year %in% c(2021, 2022, 2023, 2024)){
    #Produces error if invalid year is specified
    stop("Error: `year` is not valid. It must be 2021, 2022, 2023, or 2024")
  }
  
  #Checking to make sure the values specified for `numeric_vars` are valid
  #Defining the acceptable numeric variables
  accepted_numeric <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
                        "PWGTP")
  
  #Making sure that the numeric values inputted by user are valid
  if(!all(numeric_vars %in% accepted_numeric)){
    #Produces error if invalid numeric value is specified
    stop("Error: Invalid numeric variable specified")
  }
  
  #Making sure the PWGTP variable is included
  if(!"PWGTP" %in% numeric_vars){
    #Produces error if PWGTP variable is not included
    stop("Error: Variable 'PWGTP' must be included in `numeric_vars`")
  }
  
  # Making sure at least one other numeric variable along with PWGTP is 
  # specified. setdiff(numeric_vars, "PWGTP") returns all elements in 
  # numeric_vars that are not "PWGTP". Therefore if 
  # length(setdiff(numeric_vars,"PWGTP")) is less than 1, that means that the
  # only variable used was PWGTP
  if(length(setdiff(numeric_vars, "PWGTP")) < 1){
    #Produces error if only "PWGTP" is specified for numeric_vars
    stop("Error: At least one other numeric variable other than 'PWGTP' must 
    be specified")
  }
  
  #Checking to make sure the values specified for `cat_vars` are valid
  #Defining accepted categorical variables
  accepted_cat <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  
  #Making sure the specified categorical variable is accepted
  if(!all(cat_vars %in% accepted_cat)){
    #Produces error if invalid categorical variable is specified
    stop("Error: `cats_var` must be one of `FER`, `HHL`, `SCH`, `SCHL`, or
         `SEX`")
  }
  
  #Making sure at least one value is specified for `cat_vars`
  if(length(cat_vars) < 1){
    #Produces error if no categorical variable is specified
    stop("Error: `cats_var` must be one of `FER`, `HHL`, `SCH`, `SCHL`, or 
    `SEX`")
  }
  
  #Checking to make sure the value specified for geo_level is valid. This 
  #checks to make sure only one geographical level is specified, the user did   #not specify "All", and the user specified one of "Region", "Division", or 
  #"State"
  if(length(geo_level) != 1 || geo_level == "All" || 
     !geo_level %in% c("Region", "Division", "State")){
    #Produces error if invalid geographical level is specified
    stop("Error: `geo_level` must be one of `Region`, `Division`, or 
         `State`")
  }
  
  #Setting `geo_level` specification to the Census API variable name as the
  #state variable is `ST` for the years 2021 and 2022 and `STATE` for 2023 
  #and 2024
  if(geo_level == "State"){
    geo_param <- ifelse(year %in% c(2021,2022), "ST", "STATE")
  } else if(geo_level == "Region"){
    geo_param <- "REGION"
  } else if(geo_level == "Division"){
    geo_param <- "DIVISION"
  }
  
  #Specifying default values for the `geo_subset` variable if user provided 
  #NULL. For ease, the default subset for each geographic level will be 1
  if(is.null(geo_subset)){
    if(geo_level == "State"){
      #Using '01' as the Census State codes are two digits
      geo_subset <- "01"
    } else if(geo_level == "Region"){
      geo_subset <- "1"
    } else if(geo_level == "Division"){
      geo_subset <- "1"
    }
  }
  
  #Formatting the value specified for `geo_subset` to fit the format of the 
  #API URL. If user specified "All", it will return "*" as that is the symbol
  #that will return all states, regions, or divisions. If `geo_subset` has 1 
  #value it will return the value as a string and if it has multiple values,
  #it will return the value as a comma-separated string
  if(tolower(geo_subset[1]) == "all"){
    geo_sub_level <- "*"
  }else{
    geo_sub_level <- paste(geo_subset, collapse = ",")
  }
  
  #This part of the function will setup the url and the `GET` function.
  #The census API is 
  #https://api.census.gov/data/year/acs/acs1/pums?get=var_names&for=var
  #Therefore I need a way to automatically apply the variables for get= and
  #for=
  #Saving url for the given year specified
  year_url <- paste0("https://api.census.gov/data/", year, "/acs/acs1/pums")
  
  #Combining numeric and categorical variables for get= part of url
  combined_vars <- unique(c(numeric_vars,cat_vars))
  
  #Joining variable names and separating them by commas for get= part of url
  get_param <- paste(combined_vars, collapse = ",")
  
  #Creating the for= part of url so we get for=geo_param:geo_sub_level,
  #making sure geo_param is lowercase to fit the url structure
  for_param <- paste0(tolower(geo_param), ":", geo_sub_level)
  
  #Calling `GET` function to retrieve API data
  response <- GET(year_url, query = list(get = get_param, `for` = for_param, 
                                         key = key))
  
  #Saving response variable as a tibble
  census_tibble <- parse_api_response(response)
  
  #Reading in variable values from the `.rds` file I created earlier
  var_vals <- readRDS("data/pums_variables.rds")
  
  #Saving the variable values for the year specified
  year_vals <- var_vals[[as.character(year)]]
  
  #Now I can apply the necessary transformations for the numeric and 
  #categorical variables respectively. 
  #Starting with the numeric variables
  for (num_var in numeric_vars) {
  if (num_var %in% c("JWAP", "JWDP")) {
    
    # 1. Safely retrieve the dictionary (checking $item and $values$item)
    var_meta <- year_vals[[num_var]]
    time_dict <- if (!is.null(var_meta$item)) var_meta$item else var_meta$values$item
    
    # Guard clause: Stop if metadata couldn't be located
    if (is.null(time_dict)) {
      stop(sprintf("Metadata dictionary for '%s' was not found in year_vals. Check str(year_vals[['%s']]).", num_var, num_var))
    }
    
    # 2. Extract values and codes into a named mapping vector
    label_map <- unname(unlist(time_dict))
    names(label_map) <- as.character(names(time_dict))
    
    # 3. Map raw API character codes directly ("0", "001", etc.)
    raw_codes <- sprintf("%03d", as.numeric(census_tibble[[num_var]]))
    time_strings <- label_map[raw_codes]
    
    # 4. Use vapply() to enforce a numeric output vector of exact length 53,710
    census_tibble[[num_var]] <- vapply(
      time_strings,
      parse_time_string2,
      FUN.VALUE = numeric(1),
      USE.NAMES = FALSE
    )
    
  } else {
    census_tibble[[num_var]] <- as.numeric(census_tibble[[num_var]])
  }
}
  
  #Now for the categorical variables
  for(cat_var in cat_vars){
    #Saving the variable values from year_vals
    var_info <- year_vals[[cat_var]]$values$item
    
    #Saving the levels from cat_vars using names() as the levels are the 
    #variable names
    levels_val <- as.character(names(var_info))
    
    #Saving the level labels from cat_vars using unlist() as the labels are 
    #stored in a list
    labels_val <- unname(unlist(var_info))
    
    #Converting variable to factor with appropriate levels and labels
    census_tibble[[cat_var]] <- factor(
      as.character(census_tibble[[cat_var]]), 
      levels = levels_val,
      labels = labels_val)
  }
  
  class(census_tibble) <- c("census", class(census_tibble))
  
  #Returning the census tibble
  return(census_tibble)
}
```

```{r}
get_census_tibble(year = 2024, numeric_vars = c("JWDP","PWGTP"), cat_vars = "SCH", geo_level = "State", geo_subset = NULL)
```



```{r}
# for(num_var in numeric_vars){
#     #Checking if numeric variable is time based (JWAP or JWDP)
#     if(num_var %in% c("JWAP", "JWDP")){
#       #Mapping time codes from census_tibble to strings from year_vals using 
#       #unlist() as the strings from year_vals are in a list
#       time_codes <- unlist(year_vals[[num_var]]$item)[census_tibble[[num_var]]]
#       
#       #Converting strings to midpoint minutes
#       census_tibble[[num_var]] <- sapply(time_codes, 
#                                          parse_time_string2)
#     } else{
#       #For when num_var is not time based
#       census_tibble[[num_var]] <- as.numeric(census_tibble[[num_var]])
#     }
#   }
```