---
title: "Mason Donehoo Project 1"
format: html
editor: visual
---

```{r}
#| include: FALSE
library(httr)
library(jsonlite)
library(dplyr)
library(tidyr)
library(tibble)
```

## Task 1: Data Processing

### Parsing Variable Values

For this section, I am retrieving and parsing metadata for the Public Use Microdata Sample (PUMS) Census API across four years: 2021, 2022, 2023, and 2024. The variables we are interested in are `AGEP`, `GASP`, `GRPIP`, `JWAP`, `JWDP`, `JWMNP`, `PWGTP`, `FER`, `HHL`, `SCH`, `SCHL`, `SEX`, `ST` (for 2021, 2022)/`STATE` (for 2023, 2024), `REGION`, and `DIVISION`. The variables for each year will be stored in a list that will then be saved in a `.rds` file under the `docs` folder.

```{r}
#Creating function to fetch and parse JSON variables for a single year
parse_year_metadata <- function(year){
  #Saves the url for whatever year the user is interested in
  url <- paste0("api.census.gov/data/", year, 
                "/acs/acs1/pums/variables.json")
  
  #Using GET() to contact the website and retrieve variable specifications
  response <- GET(url)
  
  #Parsing the JSON using fromJSON(rawToChar(...), simplifyVector = F). We use simplifyVector = F to ensure we return a list 
  parsed <- fromJSON(rawToChar(response$content), simplifyVector = F)
  
  #Extracting the `variables` element from the parsed data
  vars_list <- parsed$variables
  
  #Creating a vector of the variables we are interested in
  vars_needed <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP",
                   "FER", "SCH", "SCHL", "SEX", "ST", "STATE", "REGION", 
                   "Division")
  
  #Subsetting vars_list to get only the variables we are interested in from vars_needed
}
```
